---
title: "RevisedPreprocessing"
author: "Schaler Starks"
date: "2025-05-04"
output: html_document
---
---
title: "MilkweedOccurrenceBioclim"
author: "Schaler Starks"
date: "2025-04-27"
output: html_document
---

```{r Libraries}
# R Version 4.4.2

# Install if not already
# install.packages("spocc")
# install.packages("dplyr")
#install.packages("geodata") # for climate data processing
#install.packages("terra") # for climate data processing
library(ggplot2)
library(dplyr)
library(maps)

library(spocc)
#library(raster) # unecessary if I use terra # raster::getData doesn't work, not part of R anymore
library(geodata)
library(terra)
library(sp) # generate spatial points for syriaca and speciosa occurrances
library(dismo) # for generating background points
#library(caret) # split test/train


library(leaflet)
```


```{r Load In Occurrence Data}
setwd("C:\\Users\\these\\Documents\\BIOL361\\ClimAnalysisProj2")


syriaca_inat <-occ(query = "Asclepias syriaca", from = "inat", limit = 3000)
speciosa_inat <- occ(query = "Asclepias speciosa", from = "inat", limit = 3000)
```
```{r}
# Making data frames
syriaca_df <- occ2df(syriaca_inat)
speciosa_df <- occ2df(speciosa_inat)

```

```{r}
# Check a sample of the data
head(syriaca_inat)
colnames(syriaca_inat)
dim(syriaca_inat)
```

```{r}

```

```{r Visualization from BIOCLIM_example}
occ <- bind_rows(syriaca_df, speciosa_df)

na_map <- map_data("world", region = c("USA", "Canada", "Mexico"))
occ$longitude <- as.numeric(occ$longitude)
occ$latitude <- as.numeric(occ$latitude)

occurrence_plot <- ggplot() +
  geom_polygon(data = na_map, aes(x = long, y = lat, group = group),
               fill = "gray90", color = "gray50") +
  geom_point(data = occ, aes(x = longitude, y = latitude, color = name), size = 2,
             alpha = 0.3) +
  scale_color_manual(values = c("Asclepias syriaca" = "orange", 
                                "Asclepias speciosa" = "lightblue")) +
  coord_fixed(1.3, xlim = c(-130, -60), ylim = c(20, 55)) +
  theme_minimal() +
  labs(title = "Occurrences of Asclepias syriaca and speciosa in North America",
       color = "name")
occurrence_plot
```

```{r }
occ
```

```{r Save the Visualization}
ggsave("occurrence_map.png", plot = occurrence_plot, width = 10, height = 6, dpi = 300)
```

```{r Dates for CO2 Emmissions Later}
library(lubridate)

# change to date type
speciosa_df$date <- ymd(speciosa_df$date)
syriaca_df$date <- ymd(syriaca_df$date)

# extract the year
speciosa_df$year <- year(speciosa_df$date)
syriaca_df$year <- year(syriaca_df$date)

# Combine list of years
combined_years <- sort(unique(c(speciosa_df$year, syriaca_df$year)))

print(combined_years)
```
```{r}
library(lubridate)

# change to date type
occ$date <- ymd(occ$date)
# extract the year
occ$year <- year(occ$date)
# Combine list of years
years <- sort(unique(c(occ$year)))

print(years)
```

```{r Remove Original iNaturalist to clear space}
rm(syriaca_inat, speciosa_inat)
```

## Prep Bioclim Data

```{r Climate Occurrence Data}
bioclim <- geodata::worldclim_global(var = "bio", res = 2.5, path = "data/")

# to define us bounding box with new documentation
us_extent <- terra::ext(-125, -65, 25, 50)
bioclim_us <- terra::crop(bioclim, us_extent)

head(bioclim_us)
#rm(bioclim)
```


```{r Occurrence Data with Bioclim}
# Occurrence Data: Extract bioclim for Asclepias syriaca points
occ %>%
  na.omit() %>% # this line removes the null values
  vect(geom = c("longitude", "latitude"), crs = crs(bioclim_us)) %>%
  extract(bioclim_us, .) %>%
  mutate(presence = 1)

speciosa_df %>%
  mutate(
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)
  ) %>%
  na.omit() %>% # this line removes the null values
  vect(geom = c("longitude", "latitude"), crs = crs(bioclim_us)) %>%
  extract(bioclim_us, .) %>%
  mutate(presence = 1)

speciosa_df %>%
  mutate(
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)
  ) %>%
  na.omit() %>%
  vect(geom = c("longitude", "latitude"), crs = crs(bioclim_us)) %>% # this line instead of spatial points
  extract(bioclim_us, .) %>%
  mutate(presence = 1)
```
```{r}
occ
```

Researcher's note: Here I choose to generate 16 pseudoabsences for every true presence. Liu et al. (2019) found that for rare species, the number of back-ground points needs to be up to 16 times greater than the number of true presences for model accuracy to reach an asymptote. In addition, pseudo-absence points can’t fall within 4 km of an occurrence


```{r Prepping Absence Data}
# Number of pseudo-absences = 16 × number of true presences
#n_presence <- nrow(syriaca_df) + nrow(speciosa_df)
n_presence <- nrow(occ)
n_pseudo <- n_presence * 16
```

```{r Sampling More Random Points than Needed}
set.seed(27)

# Sample background points (oversample ~2x to be safe)
bg_points <- spatSample(bioclim_us[[1]],
                        size = n_pseudo * 2,
                        method = "random",
                        na.rm = TRUE,
                        xy = TRUE)
```

```{r}
bg_vect <- vect(bg_points, 
                geom = c("x", "y"), 
                crs = crs(bioclim_us))

presence_vect <- vect(occ, geom = c("longitude", "latitude"), crs = crs(bioclim_us))
```

```{r}
# Split background points into smaller chunks
chunk_size <- 1000
n_chunks <- ceiling(nrow(bg_vect) / chunk_size)

# Store minimum distances for each background point
min_distances <- numeric(nrow(bg_vect))

for (i in 1:n_chunks) {
  index_start <- (i - 1) * chunk_size + 1
  index_end <- min(i * chunk_size, nrow(bg_vect))
  
  bg_chunk <- bg_vect[idx_start:idx_end, ]
  
  # Compute distances 
  dist_matrix <- terra::distance(bg_chunk, presence_vect)
  
  # Get the min dist to any presence point
  min_distances[idx_start:idx_end] <- apply(dist_matrix, 1, min)
}
bg_vect_far <- bg_vect[min_distances > 4000, ]
bg_vect_final <- bg_vect_far[sample(1:nrow(bg_vect_far),
                                    n_pseudo, 
                                    replace = TRUE), ]

```

### CO2 Emmissions Data
```{r}
# List all the .nc files
co2_files <- list.files("C:\\Users\\these\\Documents\\BIOL361\\ClimAnalysisProj2\\RelevantCO2",  
                        pattern = ".nc$", 
                        full.names = TRUE)

# Load them into a SpatRaster collection
co2_stack <- rast(co2_files)
```

```{r}
```


```{r}
occ_totalData <- occ |>
  mutate(presence = 1) %>%
  bind_rows(
    extract(bioclim_us, bg_vect_final) %>%
      dplyr::select(-ID) %>%
      mutate(presence = 0)
  )
```

```{r}
# Combine presence 'n' pseudoabsence spatial vector
combined_vect <- vect(occ_totalData, geom = c("longitude", "latitude"), crs = crs(co2_stack))

# Extract CO₂ values
co2_vals <- extract(co2_stack, combined_vect)

# Remove the ID column added by extract
co2_vals <- co2_vals[, -1]

# Bind CO₂ values to `occ_totalData`
occ_totalData <- bind_cols(occ_totalData, co2_vals)

```


```{r}
totalData_clean <- occ_totalData %>%
  na.omit()
```

```{r}
write.csv(totalData_clean, "milkweedData.csv", row.names = FALSE)
```

